---
title: "stack plot"
author: "Jun Xu"
date: "8 Apr., 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## input files and parameters

"-t", "--type", such as **p c o f g, for phylum class order family genus**;

"-d", "--design", the metadata such as `../design.txt`;

"-n", "--group", default as the "group" column in the "design.txt", which can be changed manually;

"-b", "--number", int., for choosing species with most highest abundance for presentation;

"-o", "--output", for define the path to save results;

"-w", "--width", figure width;

"-e", "--height", figure height.

```{r main}
options(warn = -1)
#  Clean enviroment object
rm(list=ls()) 


# 1.2 setting source
site="https://mirrors.tuna.tsinghua.edu.cn/CRAN"
# install packages, if needed
if (!suppressWarnings(suppressMessages(require("optparse", character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))) {
  install.packages(p, repos=site)
  require("optparse",character.only=T) 
}

if (TRUE){
  option_list = list(
    make_option(c("-i", "--input"), type="character", default="../data/2-1a.tax_6Genus0.001_HC.txt",
                help="Input file; phylum(p) class(c) order(o) family(f) genus(g) [default %default]"),
    make_option(c("-d", "--design"), type="character", default="../data/2-1b.HC.cuttree.txt",
                help="metadata file;  [default %default]"),
    make_option(c("-n", "--group"), type="character", default="data.cluster",
                help="name of group type;  [default %default]"),
    make_option(c("-b", "--number"), type="numeric", default=24,
                help="Number taxonomy for showing;  [default %default]"),
    make_option(c("-w", "--width"), type="numeric", default=6,
                help="Width of figure;  [default %default]"),
    make_option(c("-e", "--height"), type="numeric", default=4,
                help="Height of figure;  [default %default]"),
    make_option(c("-o", "--output"), type="character", default="2-1.stackplot", 
                help="output directory or prefix; [default %default]")
  )
  opts = parse_args(OptionParser(option_list=option_list))
  
  if (opts$input==""){opts$input=paste("tax_",opts$type,".txt", sep = "")}
  if (opts$output==""){opts$output=paste("2-1.stackplot",opts$type,"",sep = "")}
  
  # confirming the file names and path
  print(paste("Taxonomy level is ", opts$input,  sep = ""))
  print(paste("Number taxonomy for showing is ", opts$number,  sep = ""))
  print(paste("The metadata file is ", opts$design,  sep = ""))
  print(paste("The group name is ", opts$group,  sep = ""))
  print(paste("Output figure width ", opts$width,  sep = ""))
  print(paste("Output figure height ", opts$height,  sep = ""))
  print(paste("The output file prefix is ", opts$output, sep = ""))
}


# 2. install packages
# 2.1 install CRAN packages
package_list = c("reshape2","ggplot2")

for(p in package_list){
  if(!suppressWarnings(suppressMessages(require(p, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))){
    install.packages(p, repos=site)
    suppressWarnings(suppressMessages(library(p, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))
  }
}

# 2.2 install bioconductor packages
package_list = c("digest","ggrepel","RColorBrewer")
for(p in package_list){
  if(!suppressWarnings(suppressMessages(require(p, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))){
    BiocManager::install(p)
    suppressWarnings(suppressMessages(library(p, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))
  }
}

# 2.3 install Github packages
package_list = c("kassambara/ggpubr")
for(p in package_list){
  q=unlist(strsplit(p,split = "/"))[2]
  if(!suppressWarnings(suppressMessages(require(q, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))){
    install_github(p)
    suppressWarnings(suppressMessages(library(q, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)))
  }
}


```

## reading input files

```{r read}

# 3. reading files
# reading taxa file
tax_sample = read.table(opts$input, header=T, row.names= 1, sep="\t", comment.char="") 
rownames(tax_sample) = gsub("k__Bacteria\\|p__", "", rownames(tax_sample))


# reading metadata file
design = read.table(opts$design, header=T, row.names= 1, sep="\t", comment.char="") 

# defining group
sampFile = data.frame(group=design[,opts$group],
                      sample=row.names(design), 
                      row.names = row.names(design))

# data filtering
idx = rownames(sampFile) %in% colnames(tax_sample) # match design with alpha
sampFile = sampFile[idx,]
tax_sample = tax_sample[,rownames(sampFile)] 
tax_sample

```

## Visualization

```{r sample}

# 4. Statistics and Visualization
# 4.1 Stackplot for each samples
# sorting
mean_sort = tax_sample[(order(-rowSums(tax_sample))), ]
mean_sort = as.data.frame(mean_sort)

other = colSums(mean_sort[opts$number:dim(mean_sort)[1], ])
mean_sort = mean_sort[1:(opts$number - 1), ]
mean_sort = rbind(mean_sort,other)
rownames(mean_sort)[opts$number] = c("Others")

# saving data and output files, if needed, rename HC into RA
merge_tax=mean_sort
write.table("\t", file=paste(opts$output,"_sample_HC.txt",sep=""),append = F, quote = F, eol = "", row.names = F, col.names = F)
write.table(merge_tax, file=paste(opts$output,"_sample_HC.txt",sep=""), append = T, quote = F, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = T, col.names = T)

# Visualization
mean_sort$tax = rownames(mean_sort)
data_all = as.data.frame(melt(mean_sort, id.vars=c("tax")))
data_all = merge(data_all, sampFile, by.x="variable", by.y = "sample")
## to order the tax
# (1) for HC
data_all$tax <- factor(data_all$tax, levels = c("Bacteroides","Roseburia","Agathobacter","Faecalibacterium","Blautia",
                                               "Lachnoclostridium","Fusicatenibacter","Lachnospiraceae_unidentified",
                                               "Escherichia_Shigella","Bifidobacterium","Dorea","Ruminococcus_2",
                                               "Ruminococcus_torques_group","Megamonas","Ruminococcus_gnavus_group",
                                               "Lachnospira","Subdoligranulum","Dialister","Lachnospiraceae_NK4A136_group",
                                               "Streptococcus","Klebsiella","Enterococcus","Prevotella_9","Others"),ordered = T)
# # (2) for RA
# data_all$tax <- factor(data_all$tax, levels = c("Bacteroides","Roseburia","Faecalibacterium","Blautia","Agathobacter",
#                                                 "Ruminococcus_gnavus_group","Lachnoclostridium","Bifidobacterium",
#                                                 "Fusicatenibacter","Escherichia_Shigella","Lachnospiraceae_unidentified",
#                                                 "Streptococcus","Lachnospira","Ruminococcus_2","Dorea","Ruminococcus_torques_group",
#                                                 "Lachnospiraceae_NK4A136_group","Subdoligranulum","Megamonas","Dialister",
#                                                 "Clostridium_sensu_stricto_1","Klebsiella","Prevotella_9","Others"),ordered = T)

getPalette =c(brewer.pal(n = 12, name = "Paired"),brewer.pal(n = 8, name = "Set2"),brewer.pal(n = 8, name = "Set1"))

p = ggplot(data_all, aes(x=variable, y = value, fill = tax )) + 
  geom_bar(stat = "identity",position="fill", width=1)+scale_fill_manual(values = getPalette) +
  scale_y_continuous(labels = scales::percent) + 
  facet_grid( ~ group, scales = "free_x", switch = "x") +  theme(strip.background = element_blank())+
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())+
  xlab("Groups")+ylab("Percentage (%)")+ theme_classic()+theme(axis.text.x=element_text(angle=45,vjust=1, hjust=1))
p

# saving figures
ggsave(paste(opts$output, "_sample_HC.pdf", sep=""), p, width = 3*opts$width, height = opts$height)

```

## 按组均值绘制柱状图

```{r group}

# 4.2 Visualization based on groups
# Acquiring group depending average abundance
mat_t = t(merge_tax)
mat_t2 = merge(sampFile, mat_t, by="row.names")
mat_t2 = mat_t2[,c(-1,-3)]

# mean value
mat_mean = aggregate(mat_t2[,-1], by=mat_t2[1], FUN=mean) # mean
mat_mean_final = do.call(rbind, mat_mean)[-1,]
geno = mat_mean$group
colnames(mat_mean_final) = geno

# saving data and output files, if needed, rename HC into RA.
mean_sort=as.data.frame(mat_mean_final)
write.table("\t", file=paste(opts$output,"_group_HC.txt",sep=""),append = F, quote = F, eol = "", row.names = F, col.names = F)
write.table(mean_sort, file=paste(opts$output,"_group_HC.txt",sep=""), append = T, quote = F, sep="\t", eol = "\n", na = "NA", dec = ".", row.names = T, col.names = T)

# reshaping data frame
mean_sort$tax = rownames(mean_sort)
data_all = as.data.frame(melt(mean_sort, id.vars=c("tax")))

## to order the tax
# (1) for HC
data_all$tax <- factor(data_all$tax, levels = c("Bacteroides","Roseburia","Agathobacter","Faecalibacterium","Blautia",
                                               "Lachnoclostridium","Fusicatenibacter","Lachnospiraceae_unidentified",
                                               "Escherichia_Shigella","Bifidobacterium","Dorea","Ruminococcus_2",
                                               "Ruminococcus_torques_group","Megamonas","Ruminococcus_gnavus_group",
                                               "Lachnospira","Subdoligranulum","Dialister","Lachnospiraceae_NK4A136_group",
                                               "Streptococcus","Klebsiella","Enterococcus","Prevotella_9","Others"),ordered = T)
# # (2) for RA
# data_all$tax <- factor(data_all$tax, levels = c("Bacteroides","Roseburia","Faecalibacterium","Blautia","Agathobacter",
#                                                 "Ruminococcus_gnavus_group","Lachnoclostridium","Bifidobacterium",
#                                                 "Fusicatenibacter","Escherichia_Shigella","Lachnospiraceae_unidentified",
#                                                 "Streptococcus","Lachnospira","Ruminococcus_2","Dorea","Ruminococcus_torques_group",
#                                                 "Lachnospiraceae_NK4A136_group","Subdoligranulum","Megamonas","Dialister",
#                                                 "Clostridium_sensu_stricto_1","Klebsiella","Prevotella_9","Others"),ordered = T)

# Visualization
p = ggplot(data_all, aes(x=variable, y = value, fill = tax )) + 
  geom_bar(stat = "identity",position="fill", width=0.7)+ scale_fill_manual(values = getPalette)+
  scale_y_continuous(labels = scales::percent) + 
  xlab("Groups")+ylab("Percentage (%)")+ theme_classic()
p

# saving figures
ggsave(paste(opts$output, "_group_HC.pdf", sep=""), p, width = opts$width, height = opts$height)

```


If used this script, please cited:
  Yun Liu, Jun Xu, Xinhua Ren, Yu Zhang, Ziliang Ke, Jianhua Zhou, Yang Wang, Yifan Zhang & Yulan Liu (2022) Cholecystectomy-induced secondary bile acids accumulation ameliorates colitis through inhibiting monocyte/macrophage recruitment, Gut Microbes, 14:1, 2107387, DOI: 10.1080/19490976.2022.2107387

Copyright 2024-2026 Jun Xu <xujun@hsc.pku.edu.cn>